Follow-up Samples

### 2017-02-22 submission/manuscript_outline_20161024.Rmd

**Results:** The bacterial microbiome changed more in response to lesion removal in carcinoma cases compared to adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).  There was no difference for either the adenoma or carcinoma group in the relative abundance of any OTU between the pre and post lesion removal (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).  A model built to classify lesion had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while a model built to classify initial versus follow up samples had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.  Follow up samples had a decrease in the positive probability of lesion (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05) suggesting a movement towards a more normal bacterial community.  The initial sample model had a decrease in positive probability for the follow up samples tobe an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).  The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.  A total of `r length(common_vars_summary$otu)` OTUs were common to both models with the majority of these classifying to commensal bacteria (e.g. *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*).

@@

Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is currently the third most common cause of cancer deaths [@jemal_cancer_2010; @haggar_colorectal_2009].  Over the last few years rate of disease mortality has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].  However, despite this improvement there are still approximately 50,000 deaths from the disease per year [@haggar_colorectal_2009].

Over the last few years there has been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  A few studies have also shown how this bacterial microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.  These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.  Building off of these findings, there has been promising work on the bacterial microbiome and its ability to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].  Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome response to lesion (adenoma or carcinoma) removal.

In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) removal of lesion (adenoma or carcinoma).  We analyzed changes in alpha and beta diversity as well as specific Operational Taxonomic Units (OTUs) relative abundance.  We utilized Random Forest to build two models.  The first model was built to classify either lesion and normal samples while the second was built to classify initial and follow up samples.  Subsequent observations on how these models, as well as specific OTUs within them, performed pre and post lesion removal helped inform us as to whether initial and follow up samples were changing and whether it was towards a more normal bacterial microbiome or not.  We then investigated the two models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normal samples but also initial and follow up.

@@

***Bacterial Community and FIT:***
Within our `r length(rownames(good_metaf))` person cohort we first wanted to test whether there were any broad differences between initial and follow up samples based on lesion being either adenoma (n = `r length(rownames(filter(good_metaf, Diagnosis == "adenoma")))`) or carcinoma (n = `r length(rownames(filter(good_metaf, Diagnosis != "adenoma")))`).  We found that the bacterial community in those with carcinoma were more dissimilar to their initial sample than those with adenoma (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.001, paste("<"), paste(">"))` 0.001) [Figure 1a]. We also found that there were larger changes in fecal blood (measured by FIT) for those with carcinoma versus adenoma (P-value `r ifelse(change_summary['fit', 'Pvalue'] < 0.0001, paste("<"), paste(">"))` 0.0001) [Figure 1b].  The bacterial community structure before and after surgery was visualized using NMDS for both adenoma [Figure1c] (PERMANOVA `r ifelse(beta_summary['Adenoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05) and carcinoma [Figure 1d] (PERMANOVA `r ifelse(beta_summary['Carcinoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).  Interestingly, when initial and follow up samples were compared, regardless of whether the lesions were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary['All', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).  When investigating alpha diversity metrics there was no difference found between initial and follow up samples for lesion, adenoma only, or carcinoma only for any metric tested [Table S1].  We also observed that there was no difference in the relative abundance of any OTU between initial and follow up samples for lesion, adenoma only, or carcinoma only [Figure S1].

@@

***Treatment Differences:***
After observing these changes in the bacterial community and positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.  There was only a significant difference for change in positive probability for those treated with chemotherapy for the initial sample model (P-value `r ifelse(chemo_rad_summary['chemo_red_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).  This suggests that those treated for chemotherapy my have had a larger change from the initial sample then those without such treatment.  All other variables that were tested showed no difference based on whether chemotherapy or radiation was used [Table S3].

@@

Within our study there was a significant difference for the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) than carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).  These results would indicate that the findings described were specific to the surgical intervention and that some of the differences observed between carcinoma and adenoma samples could be due to differences in collection time between samples for the two different groups.  Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1a & 1d]. This confounding though would not affect the observations where these individuals were grouped together.

@@

Many of the common OTUs between the two models used had many OTUs that taxonomically classified to potential butyrate producers [Table S2].  Another batch of OTUs classified to bacteria that can either degrade polyphenols or are inhibited by them.  Both butyrate and polyphenols are thought to be protective against cancer in part by reducing inflammation [@louis_gut_2014].   These protective compounds are derived from the breakdown of fiber, fruits, and vegetables by resident gut microbes.  One example of this potential diet-microbiome-inflammation-polyp axis is that *Bacteroides*, which was highly prevalent in our models, are known to be increased in those with high non-meat based protein consumption [@zhu_intake_2016].  High protein consumption in general has been linked with an increased CRC risk [@mu_colonic_2016].  Conversely, *Bacteroides* are inhibited by polyphenols which are derived from fruits and vegetables [@ozdal_reciprocal_2016].  Our data fits with the hypothesis that the microbial metabolites from breakdown products within our own diet could not only help to shape the existing community but also have an effect on CRC risk and disease progression.

>>>>

**Results:** The bacterial microbiome changed more in response to lesion removal in carcinoma cases compared to adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).  There was no difference for either the adenoma or carcinoma group in the relative abundance of any OTU between pre and post lesion removal (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).  A model built to classify lesion had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while a model built to classify initial samples had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.  Follow up samples had a decrease in the positive probability of lesion (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05) suggesting a movement towards a more normal bacterial community.  The initial sample model had a decrease in positive probability for the follow up samples to be an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).  The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.  A total of `r length(common_vars_summary$otu)` OTUs were common to both models with the majority of these classifying to commensal bacteria (e.g. *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*).

@@

Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is currently the third most common cause of cancer deaths [@jemal_cancer_2010; @haggar_colorectal_2009].  Over the last few years the rate of disease mortality has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].  However, despite this improvement there are still approximately 50,000 deaths from the disease per year [@haggar_colorectal_2009].

Over the last few years a few studies have shown how this bacterial microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.  There has also been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.  Building off of these findings, there has been promising work on the bacterial microbiome and its ability to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].  Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome response to lesion (adenoma or carcinoma) removal.

In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) removal of lesion (adenoma or carcinoma).  We analyzed changes in alpha and beta diversity as well as specific Operational Taxonomic Units (OTUs) relative abundance.  We utilized Random Forest to build two models.  The first model was built to classify lesion and normal samples while the second was built to classify initial and follow up samples.  Subsequent observations on how these models, as well as specific OTUs within them, performed pre and post lesion removal helped inform us as to whether initial and follow up samples were changing and whether it was towards a more normal bacterial microbiome or not.  We also investigated the two models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normal samples but also initial and follow up.

@@

***Bacterial Community and FIT:***
Within our `r length(rownames(good_metaf))` person cohort we first wanted to test whether there were any broad differences between initial and follow up samples based on lesion being either adenoma (n = `r length(rownames(filter(good_metaf, Diagnosis == "adenoma")))`) or carcinoma (n = `r length(rownames(filter(good_metaf, Diagnosis != "adenoma")))`).  We found that the bacterial community in those with carcinoma were more dissimilar to their initial sample than those with adenoma (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.001, paste("<"), paste(">"))` 0.001) [Figure 1A]. We also found that there were larger changes in fecal blood (measured by FIT) for those with carcinoma versus adenoma (P-value `r ifelse(change_summary['fit', 'Pvalue'] < 0.0001, paste("<"), paste(">"))` 0.0001) [Figure 1B].  The bacterial community structure before and after surgery was visualized using NMDS for both adenoma [Figure 1C] (PERMANOVA `r ifelse(beta_summary['Adenoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05) and carcinoma [Figure 1D] (PERMANOVA `r ifelse(beta_summary['Carcinoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).  Interestingly, when initial and follow up samples were compared, regardless of whether the lesions were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary['All', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).  When investigating alpha diversity metrics there was no difference found between initial and follow up samples for lesion, adenoma only, or carcinoma only for any metric tested [Table S1].  We also observed that there was no difference in the relative abundance of any OTU between initial and follow up samples for lesion, adenoma only, or carcinoma only [Figure S1].

@@

***Treatment Differences:***
After observing these changes in the bacterial community and positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.  There was only a significant difference for change in positive probability for those treated with chemotherapy for the initial sample model (P-value `r ifelse(chemo_rad_summary['chemo_red_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).  This suggests that follow up samples for those treated for chemotherapy my have had a larger change from the initial sample then those without such treatment.  All other variables that were tested showed no difference based on whether chemotherapy or radiation was used [Table S3].

@@

Within our study there was a significant difference for the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) than carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).  These results would indicate that the findings described were specific to the surgical intervention and that some of the differences observed between carcinoma and adenoma samples could be due to differences in collection time between samples for the two different groups.  Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1A & 1D]. This confounding though would not affect the observations where these individuals were grouped together [Figure 3 & 4].

@@

Many of the common OTUs between the two models had OTUs that taxonomically classified to potential butyrate producers [Table S2].  Another batch of OTUs classified to bacteria that can either degrade polyphenols or are inhibited by them.  Both butyrate and polyphenols are thought to be protective against cancer in part by reducing inflammation [@louis_gut_2014].   These protective compounds are derived from the breakdown of fiber, fruits, and vegetables by resident gut microbes.  One example of this potential diet-microbiome-inflammation-polyp axis is that *Bacteroides*, which was highly prevalent in our models, are known to be increased in those with high non-meat based protein consumption [@zhu_intake_2016].  High protein consumption in general has been linked with an increased CRC risk [@mu_colonic_2016].  Conversely, *Bacteroides* are inhibited by polyphenols which are derived from fruits and vegetables [@ozdal_reciprocal_2016].  Our data fits with the hypothesis that the microbial metabolites from breakdown products within our own diet could not only help to shape the existing community but also have an effect on CRC risk and disease progression.
